Human and Economic Costs of Weather Events: A United States View
========================================================
The purpose of this analysis is to measure the costs human life and ecomonic costs of significant weather events.  The data for this analysis comes from the NOAA storm database, with measured storm characteristics across the United States, from 1950 - 2011.  This report will use a raw data file, perform the necessary conversions for any subsequent analysis, and present the results.  
## Data Processing 

First we must download the necessary data and reduce it's size.  The initial data, once decompressed contains 902,297 observations  of individual storms with 37 characteristic variables. 
```{r}
setwd("C:/coursera/RepData_PeerAssessment2")
# retrieving the data
if(!file.exists("NOAA.csv.bz2")){
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, "NOAA.csv.bz2")}

#load data, and decompress
d <- read.csv(bzfile("NOAA.csv.bz2"))
name <- c("BGN_DATE", "STATE", "EVTYPE", "FATALITIES", 
          "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
data <- d[,(names(d) %in% name)]
names(data) <- tolower(names(data))
```

Next, the data needs to be processed appropriately. First looking at the economic impact variables, property and crop damage are described by two variables each, first a numeric component and second a magnitude as thousands (k), millions (m), and billions (b).  The following code chunk combines these two into one dollar figure.  
```{r}
#Adjusting property and crop damage costs into appropriate numeric values
# Property damage
data$propdmg[as.character(data$propdmgexp) %in% c("k", "K")] <- 
        data$propdmg[as.character(data$propdmgexp) %in% c("k", "K")] * 1000
data$propdmg[as.character(data$propdmgexp) %in% c("m", "M")] <- 
        data$propdmg[as.character(data$propdmgexp) %in% c("m", "M")] * 1000000
data$propdmg[as.character(data$propdmgexp) %in% c("b", "B")] <- 
        data$propdmg[as.character(data$propdmgexp) %in% c("b", "B")] * 1000000000
# Crop damage
data$cropdmg[as.character(data$cropdmgexp) %in% c("k", "K")] <- 
        data$cropdmg[as.character(data$cropdmgexp) %in% c("k", "K")] * 1000
data$cropdmg[as.character(data$cropdmgexp) %in% c("m", "M")] <- 
        data$cropdmg[as.character(data$cropdmgexp) %in% c("m", "M")] * 1000000
data$cropdmg[as.character(data$cropdmgexp) %in% c("b", "B")] <- 
        data$cropdmg[as.character(data$cropdmgexp) %in% c("b", "B")] * 1000000000
```
Next the events are described in a messy fashion, with misspelled and both lower and upper case factor levels.  We attempt to consolidate redundent levels in the next code chunk, and for the sake of brevity, any events that describe multiple events, the first description is used.  

The following process eliminates some level of detail, but it will improve precision by collapsing many closely related groups into individual factor units.  This will help to provide a more general picutre of the overall costs of different events, and once this is determined one can go back and drill into the details of any individual event.  

```{r}
data$evtype <- tolower(data$evtype)
tempSep <- strsplit(data$evtype, "[/;,.-\\]")

#function to use the first discriptor before seperator
first.elm <- function(x){x[1]} #simple extractor function
data$evtype <- unlist(sapply(tempSep, first.elm))

#the and seperator, and dropping any trailing and
andSep <- strsplit(data$evtype, "( and | & )")
data$evtype <- unlist(sapply(andSep, first.elm))
data$evtype <- gsub(" [Aa][Nn][Dd]$", "", data$evtype )

#eliminate trailing numbers, for simplification of evaluation
data$evtype <- gsub("[0-9]*$", "", data$evtype) #trailing numbers
data$evtype <- gsub(" .g[0-9]*.$", "", data$evtype) #MPH numbers

#errant puncuation and letters
data$evtype <- gsub(" \\(.*$", "", data$evtype) #for parenthesis with char
data$evtype <- gsub("\\($", "", data$evtype) #for parenthesis at end
data$evtype <- gsub("-", " ", data$evtype)
data$evtype <- gsub(" +[A-Za-z] +", "", data$evtype) #middle single letters
data$evtype <- gsub(" +[A-Za-z]$", "", data$evtype) #trailing single letters

#consolidating pluarl and erronous descriptors
data$evtype <- gsub("ss$", "s", data$evtype) #double plural s i.e. ...ss
data$evtype <- gsub("drynes" ,"dry" , data$evtype)        
data$evtype <- gsub("wetnes(s?)" ,"wet" , data$evtype)
data$evtype <- gsub("ies$", "y", data$evtype)
data$evtype <- gsub("floes", "flow", data$evtype)
data$evtype <- gsub("oes", "o", data$evtype)
data$evtype <- gsub("s$", "", data$evtype) #to eliminate remaining plurality

#abbreviations
data$evtype <- gsub("tstm", "thunder storm", data$evtype)
data$evtype <- gsub("hvy", "heavy", data$evtype)
data$evtype <- gsub("wnd", "wind", data$evtype)
data$evtype <- gsub("hvy", "heavy", data$evtype)
data$evtype <- gsub("sml", "small", data$evtype)
data$evtype <- gsub("fld", "flood", data$evtype)
data$evtype <- gsub("precip$", "precipitation", data$evtype )

#lots of storm misspelling
data$evtype <- gsub("storm", " storm", data$evtype)
data$evtype <- gsub("storm[ws]", "storm", data$evtype)
data$evtype <- gsub("thundertorm", "thunder storm", data$evtype)
data$evtype <- gsub("thundestorm", "thunder storm", data$evtype)
data$evtype <- gsub("thunerstorm", "thunder storm", data$evtype)
data$evtype <- gsub("tunderstorm", "thunder storm", data$evtype)
data$evtype <- gsub("thuder", "thunder", data$evtype)
data$evtype <- gsub("thundeer", "thunder", data$evtype)
data$evtype <- gsub(" win$", " wind", data$evtype)
data$evtype <- gsub("stormind", "storm wind", data$evtype)
data$evtype <- gsub("thuner", "thunder", data$evtype)
data$evtype <- gsub("thundere", "thunder", data$evtype)
data$evtype <- gsub("winds", "wind", data$evtype)
data$evtype <- gsub("windhail", "wind hail", data$evtype)
data$evtype <- gsub("sorm", "storm", data$evtype)
data$evtype <- gsub("strom", "Storm", data$evtype)


#tropical storm and hurricane names
data$evtype <- gsub("hurricane .*$", "hurricane", data$evtype)
data$evtype <- gsub("tropical storm .*$", "tropical storm", data$evtype)

#date summaries, consolodated into one factor
data$evtype <- gsub("^summary .+$", "summary", data$evtype)

#correcting for idiosyncracies in the descriptors
data$evtype <- gsub("chill .*$" , "chill" , data$evtype) #chill and chill temperatures appears 
data$evtype <- gsub("snow-.*$" , "snow" , data$evtype)
data$evtype <- gsub("drynes" ,"dry" , data$evtype)        
data$evtype <- gsub("wetnes(s?)" ,"wet" , data$evtype)
data$evtype <- gsub("avalance" ,"avalanche" , data$evtype)  
data$evtype <- gsub("erosin", "erosion", data$evtype)
data$evtype <- gsub("coastal^ ", "coastal ", data$evtype)
data$evtype <- gsub("cstl", "coastal", data$evtype)
data$evtype <- gsub("mirco", "micro", data$evtype)
data$evtype <- gsub("devel", "devil", data$evtype)
data$evtype <- gsub("ooo", "oo", data$evtype)
data$evtype <- gsub("flood flood", "flood", data$evtype)
data$evtype <- gsub("flooding", "flood", data$evtype)
data$evtype <- gsub("precipatation", "precipitation", data$evtype)
data$evtype <- gsub("andblowing.*$", "", data$evtype)
data$evtype <- gsub("lighting", "lightning", data$evtype)
data$evtype <- gsub("w ind", "wind", data$evtype)
data$evtype <- gsub("debri", "debris", data$evtype)
data$evtype <- gsub("able", "ably", data$evtype)
data$evtype <- gsub("unseasonal", "unseasonably", data$evtype)
data$evtype <- gsub("unusual", "unusually", data$evtype)
data$evtype <- gsub("vog", "fog", data$evtype)
data$evtype <- gsub("spout", " spout", data$evtype)
data$evtype <- gsub("fire", " fire", data$evtype)
data$evtype <- gsub("flood", " flood", data$evtype)

#eliminate double spaces and white space
data$evtype <- gsub(" +( *)", " ", data$evtype) #double or multiple space
data$evtype <- gsub("^( *)", "", data$evtype) #left white space
data$evtype <- gsub("( *)$", "", data$evtype) #right white space

levels <- levels(as.factor(data$evtype))
levels <- data.frame(levels, grepl(".(s|es)$", levels))

#
```
## Results


```{r}
require(doBy)
agg.mean <- summaryBy(FATALITIES + INJURIES + PROPDMG + CROPDMG ~ EVTYPE, 
                 data = data, FUN = mean )
agg.mean <- agg.mean(order(agg.mean()))
```


##