Human and Economic Costs of Weather Events: A United States View
========================================================
The purpose of this analysis is to measure the costs human life and ecomonic costs of significant weather events.  The data for this analysis comes from the NOAA storm database, with measured storm characteristics across the United States, from 1950 - 2011.  This report will use a raw data file, perform the necessary conversions for any subsequent analysis, and present the results.  
## Data Processing 

First we must download the necessary data and reduce it's size.  The initial data, once decompressed contains 902,297 observations  of individual storms with 37 characteristic variables. 
```{r, cache = TRUE}
setwd("C:/coursera/RepData_PeerAssessment2")
# retrieving the data
if(!file.exists("NOAA.csv.bz2")){
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, "NOAA.csv.bz2")}

#load data, and decompress
data <- read.csv(bzfile("NOAA.csv.bz2"))
name <- c("BGN_DATE", "STATE", "EVTYPE", "FATALITIES", 
          "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
data <- data[,(names(data) %in% name)]
names(data) <- tolower(names(data))
```

Next, the data needs to be processed appropriately. First looking at the economic impact variables, property and crop damage are described by two variables each, first a numeric component and second a magnitude as thousands (k), millions (m), and billions (b).  The following code chunk combines these two into one dollar figure.  
```{r}
#Adjusting property and crop damage costs into appropriate numeric values
# Property damage
data$propdmg[as.character(data$propdmgexp) %in% c("k", "K")] <- 
        data$propdmg[as.character(data$propdmgexp) %in% c("k", "K")] * 1000
data$propdmg[as.character(data$propdmgexp) %in% c("m", "M")] <- 
        data$propdmg[as.character(data$propdmgexp) %in% c("m", "M")] * 1000000
data$propdmg[as.character(data$propdmgexp) %in% c("b", "B")] <- 
        data$propdmg[as.character(data$propdmgexp) %in% c("b", "B")] * 1000000000
# Crop damage
data$cropdmg[as.character(data$cropdmgexp) %in% c("k", "K")] <- 
        data$cropdmg[as.character(data$cropdmgexp) %in% c("k", "K")] * 1000
data$cropdmg[as.character(data$cropdmgexp) %in% c("m", "M")] <- 
        data$cropdmg[as.character(data$cropdmgexp) %in% c("m", "M")] * 1000000
data$cropdmg[as.character(data$cropdmgexp) %in% c("b", "B")] <- 
        data$cropdmg[as.character(data$cropdmgexp) %in% c("b", "B")] * 1000000000
```
Next the events are described in a messy fashion, with misspelled and both lower and upper case factor levels.  We attempt to consolidate redundent levels in the next code chunk, and for the sake of brevity, any events that describe multiple events, the first description is used.  

The following process eliminates some level of detail, but it will improve precision by collapsing many closely related groups into individual factor units.  This will help to provide a more general picutre of the overall costs of different events, and once this is determined one can go back and drill into the details of any individual event.  This lowers the factor levels from over 900 to 400 different factor levels, and will assist in simplifying general results

```{r, cache = TRUE}
data$evtype <- tolower(data$evtype)
tempSep <- strsplit(data$evtype, "[/;,.-\\]")

#function to use the first discriptor before seperator
first.elm <- function(x){x[1]} #simple extractor function
data$evtype <- unlist(sapply(tempSep, first.elm))

#the and seperator, and dropping any trailing and
andSep <- strsplit(data$evtype, "( and | & )")
data$evtype <- unlist(sapply(andSep, first.elm))
data$evtype <- gsub(" [Aa][Nn][Dd]$", "", data$evtype )

#eliminate trailing numbers, for simplification of evaluation
data$evtype <- gsub("[0-9]*$", "", data$evtype) #trailing numbers
data$evtype <- gsub(" .g[0-9]*.$", "", data$evtype) #MPH numbers

#errant puncuation and letters
data$evtype <- gsub(" \\(.*$", "", data$evtype) #for parenthesis with char
data$evtype <- gsub("\\($", "", data$evtype) #for parenthesis at end
data$evtype <- gsub("-", " ", data$evtype)
data$evtype <- gsub(" +[A-Za-z] +", "", data$evtype) #middle single letters
data$evtype <- gsub(" +[A-Za-z]$", "", data$evtype) #trailing single letters

#consolidating pluarl and erronous descriptors
data$evtype <- gsub("ss$", "s", data$evtype) #double plural s i.e. ...ss
data$evtype <- gsub("drynes" ,"dry" , data$evtype)        
data$evtype <- gsub("wetnes(s?)" ,"wet" , data$evtype)
data$evtype <- gsub("ies$", "y", data$evtype)
data$evtype <- gsub("floes", "flow", data$evtype)
data$evtype <- gsub("oes", "o", data$evtype)
data$evtype <- gsub("s$", "", data$evtype) #to eliminate remaining plurality

#abbreviations
data$evtype <- gsub("tstm", "thunder storm", data$evtype)
data$evtype <- gsub("hvy", "heavy", data$evtype)
data$evtype <- gsub("wnd", "wind", data$evtype)
data$evtype <- gsub("hvy", "heavy", data$evtype)
data$evtype <- gsub("sml", "small", data$evtype)
data$evtype <- gsub("fld", "flood", data$evtype)
data$evtype <- gsub("precip$", "precipitation", data$evtype )

#lots of storm misspelling
data$evtype <- gsub("storm", " storm", data$evtype)
data$evtype <- gsub("storm[ws]", "storm", data$evtype)
data$evtype <- gsub("thundertorm", "thunder storm", data$evtype)
data$evtype <- gsub("thundestorm", "thunder storm", data$evtype)
data$evtype <- gsub("thunerstorm", "thunder storm", data$evtype)
data$evtype <- gsub("tunderstorm", "thunder storm", data$evtype)
data$evtype <- gsub("thuder", "thunder", data$evtype)
data$evtype <- gsub("thundeer", "thunder", data$evtype)
data$evtype <- gsub(" win$", " wind", data$evtype)
data$evtype <- gsub("stormind", "storm wind", data$evtype)
data$evtype <- gsub("thuner", "thunder", data$evtype)
data$evtype <- gsub("thundere", "thunder", data$evtype)
data$evtype <- gsub("winds", "wind", data$evtype)
data$evtype <- gsub("windhail", "wind hail", data$evtype)
data$evtype <- gsub("sorm", "storm", data$evtype)
data$evtype <- gsub("strom", "Storm", data$evtype)

#date summaries, consolodated into one factor
data$evtype <- gsub("^summary .+$", "summary", data$evtype)

#correcting for idiosyncracies in the descriptors
data$evtype <- gsub("chill .*$" , "chill" , data$evtype) #chill and chill temperatures appears 
data$evtype <- gsub("snow-.*$" , "snow" , data$evtype)
data$evtype <- gsub("drynes" ,"dry" , data$evtype)        
data$evtype <- gsub("wetnes(s?)" ,"wet" , data$evtype)
data$evtype <- gsub("avalance" ,"avalanche" , data$evtype)  
data$evtype <- gsub("erosin", "erosion", data$evtype)
data$evtype <- gsub("coastal^ ", "coastal ", data$evtype)
data$evtype <- gsub("cstl", "coastal", data$evtype)
data$evtype <- gsub("mirco", "micro", data$evtype)
data$evtype <- gsub("devel", "devil", data$evtype)
data$evtype <- gsub("ooo", "oo", data$evtype)
data$evtype <- gsub("flood flood", "flood", data$evtype)
data$evtype <- gsub("flooding", "flood", data$evtype)
data$evtype <- gsub("precipatation", "precipitation", data$evtype)
data$evtype <- gsub("andblowing.*$", "", data$evtype)
data$evtype <- gsub("lighting", "lightning", data$evtype)
data$evtype <- gsub("w ind", "wind", data$evtype)
data$evtype <- gsub("debri", "debris", data$evtype)
data$evtype <- gsub("able", "ably", data$evtype)
data$evtype <- gsub("unseasonal", "unseasonably", data$evtype)
data$evtype <- gsub("unusual", "unusually", data$evtype)
data$evtype <- gsub("vog", "fog", data$evtype)
data$evtype <- gsub("spout", " spout", data$evtype)
data$evtype <- gsub("fire", " fire", data$evtype)
data$evtype <- gsub("flood", " flood", data$evtype)

#eliminate double spaces and white space
data$evtype <- gsub(" +( *)", " ", data$evtype) #double or multiple space
data$evtype <- gsub("^( *)", "", data$evtype) #left white space
data$evtype <- gsub("( *)$", "", data$evtype) #right white space

#tropical storm and hurricane names
data$evtype <- gsub("hurricane .*$", "hurricane", data$evtype)
data$evtype <- gsub("tropical storm .*$", "tropical storm", data$evtype)

levels <- levels(as.factor(data$evtype))
levels <- data.frame(levels, grepl(".(s|es)$", levels))
#
```
## Results
Now that we have parsed the data and cleaned it to contain the appropriate variables, lets conduct some analysis.  The code chunk below presents some summary statistics for the top ten costliest events for each of the four variables.  

```{r}
require(doBy)

#average costs
agg.mean <- summaryBy(fatalities + injuries + propdmg + cropdmg ~ evtype, 
                 data = data, FUN = mean)
#for fatalities
agg.mean <- agg.mean[order(agg.mean$fatalities.mean, decreasing = TRUE),] 
agg.mean[1:10, c(1,2)]
#for injuries
agg.mean <- agg.mean[order(agg.mean$injuries.mean, decreasing = TRUE),] 
agg.mean[1:10, c(1,3)]
#property damage
agg.mean <- agg.mean[order(agg.mean$propdmg.mean, decreasing = TRUE),] 
agg.mean[1:10, c(1,4)]
#crop damage
agg.mean <- agg.mean[order(agg.mean$cropdmg.mean, decreasing = TRUE),] 
agg.mean[1:10, c(1,5)]

#Median costs
agg.median <- summaryBy(fatalities + injuries + propdmg + cropdmg ~ evtype, 
                 data = data, FUN = median )
#for fatalities
agg.median <- agg.median[order(agg.median$fatalities.median, decreasing = TRUE),] 
agg.median[1:10, c(1,2)]
#for injuries
agg.median <- agg.median[order(agg.median$injuries.median, decreasing = TRUE),] 
agg.median[1:10, c(1,3)]
#property damage
agg.median <- agg.median[order(agg.median$propdmg.median, decreasing = TRUE),] 
agg.median[1:10, c(1,4)]
#crop damage
agg.median <- agg.median[order(agg.median$cropdmg.median, decreasing = TRUE),] 
agg.median[1:10, c(1,5)]

#aggregated total costs
agg.sum <- summaryBy(fatalities + injuries + propdmg + cropdmg ~ evtype, 
                 data = data, FUN = sum )
#for fatalities
agg.sum <- agg.sum[order(agg.sum$fatalities.sum, decreasing = TRUE),] 
agg.sum[1:10, c(1,2)]
#for injuries
agg.sum <- agg.sum[order(agg.sum$injuries.sum, decreasing = TRUE),] 
agg.sum[1:10, c(1,3)]
#property damage
agg.sum <- agg.sum[order(agg.sum$propdmg.sum, decreasing = TRUE),] 
agg.sum[1:10, c(1,4)]
#crop damage
agg.sum <- agg.sum[order(agg.sum$cropdmg.sum, decreasing = TRUE),] 
agg.sum[1:10, c(1,5)]
```
### Population Health Impacts
Looking at the human life impact statistics, the notabile results is that typically the most dangerous events are the result of either hard to predict or typically unexpected weather events, or more day to day weather events.  These include maritime accidents and tsunami's.  
Also notable is the fact that extreme heat and heatwaves seem to be relatively dangerous to population health.  The highest mean and median numbers for fatalities and injuries caused by weather events are heat related, and furthermore, the summed number deaths and injuries appears as in the top 4 most dangerous events as well.  This may be due to the lack of proper education of how to deal with heat, and how relatively easy it is to get sick in even moderate heat.  
To look at this closer, see a histogram of the heat events:
```{r}
health.hist <- data[grep("heat", data$evtype), 
                    c("evtype", "fatalities", "injuries")]

par(mfrow = c(1,2))
hist(log(health.hist$fatalities), xlab = "Fatalities")
hist(log(health.hist$injuries), xlab = "Injuries")
```
From this picture, after log transformation of the variables to aid in visual inspection, deaths related to any one heat related event are typically low, and individuals are much more likely to sustain injuries than death.

### Economic Impacts
The economic front shows a more intuitive picture of what one would expect.  Costly events are for property damage are typically large storms, such as hurricanes, typhoons, floods and fire.  For crop damage the events that seem to drive costs are related to freezing conditions, hot conditions, or dry conditions.  To visualize this, lets consider some histograms.

```{r}
#selecting top contributing events
econ.prop.hist <- data[grep("(hurricane|tropical|typhoon|flood)",data$evtype),
                       c("evtype", "propdmg")]
econ.crop.hist <- data[grep("(freez|heat|flood|drought|frost|cold|wet)", 
                            data$evtype), c("evtype", "cropdmg")]
par(mfrow = c(1,2))
hist(log(econ.prop.hist$propdmg), main = "Log of Property Damage",
     xlab = "")
hist(log(econ.crop.hist$cropdmg), main = "Log of Crop Damage", 
     xlab = "")
#for frame of reference, exponentiating 
exp(5);exp(10); exp(15)
```
Here the distribution of costs is much less skewed than the health related events previously described.  Many of these events are hard to plan for, and hard to avoid.  
